---
title: "Tutorial using GLEANR core functionality"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial using GLEANR core functionality}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(gleanr)
```

TODO: need to update the file paths to refer to `ext`, as describedin https://r-pkgs.org/vignettes.html
For now, just getting things rolling
Here we provide a single simple example of using `gleanr` on simulated data. While this doesn't comprehensively explore all of gleanr's options and functionality, we focus on the core components that will be helpful for most users.
As input, `gleanr` requires:
- A matrix (B) of GWAS effect size estimates (SNPs x traits)
- A matrix (S) of GWAS standard error estimates (SNPs x traits, same order as B)
- A matrix (C) of estimated correlation due to sample sharing between GWAS (necessary if you wish to adjust for the impact of sample sharing; traits x traits)
- A matrix (C_se) of standard error estimates corresponding to values in C (*optional* but recommended) 
- A list of variances across GWAS z-scores (*optional* to adjust correlation estimates for trait heritability)


Note that `gleanr` can be run directly from the command line using input file paths (see the `gleanr_run.R` script in the [gleanr_workflow](https://github.com/aomdahl/gleanr_workflow/blob/main/src/gleaner_run.R) repository). *We recommend this for most use cases as the simplest way to run gleanr.*

However, if users prefer to perform `gleanr` directly in R, users have two options for loading into memory necessary input files. They may:
1) Read in files manually for input to `gleanr` (allows for easier custom user data filtering) or
2) Specify file paths and use existing `gleanr` functions to tidy and check input data. *We recommend this option*


## Option 1) Manually read files in
```{r}
#fname <- system.file("shape/nc.shp", package="sf")
fpath <- "/scratch16/abattle4/ashton/snp_networks/gleanr/inst/extdata/"
library(data.table)
```




